Gravitational Lens Recovery with \Glass: How to measure the mass profile and shape of a lens

### 2014-01-21 limits/ms.tex
Minor text changes and improved Figures 1,3.

In this paper, we use a new program --- \Glass\ --- to determine what
combination of lensing, stellar mass, and stellar kinematic
constraints best constrain the projected mass profile and shape of a
gravitational lens.  The compute-intensive core of \Glass\ is a
parallel implementation of the high-dimensional sampling algorithm
from \cite{2012MNRAS.425.3077L}.  Around that core, \Glass\ has
extensible code (in software terminology, a framework) in Python,
encoding the lens equation, a basis and priors for the mass
distribution, and various post-processing modules.  \Glass\ is
described in \secref{sec:glass}.  In \secref{sec:theory} before that
we review the elements of lensing theory, stellar population
synthesis, and stellar dynamics we will need.

@@

We follow \cite{1986ApJ...310..568B} with minor differences in notation
notation.

The lens equation
\begin{equation}
\begin{aligned}
\vec\beta &= \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta) \\
\vec\alpha(\vec\theta) &= \frac{4G}{c^2D_L} \int \Sigma(\vec\theta')
                          \frac{(\vec\theta - \vec\theta')}
                          {\ |\vec\theta - \vec\theta'|^2} \, d^2\vec\theta'
\end{aligned}
\label{eqn:lens_equation}
\end{equation}
maps an observed image position $\vec\theta$ to a source position
$\vec\beta$.  The lens can then be thought of as a projected surface
density $\Sigma$ which diverts the path of a photon instantaneously
through the bending angle $\vec\alpha$.  The $D$ factors are angular
diameter distances, as in the previous section.
\begin{equation}
D_{LS} = \frac c{H_0} \frac{1+z_S}{1+z_L} \int_{z_L}^{z_S}
                      \frac{dz}{\sqrt{\Omega_m(1+z)^3 + \Omega_\Lambda}}
\end{equation}
One way to understand the lens equation is via Fermat's principle. We
can think of light as travelling only along extremum paths where
lensed images occur. Such paths occur at
the extrema of the photon {\it arrival time\/} $t(\vec\theta)$
that depends on the geometric path the photon takes and the general
relativistic gravitational time dilation due to a thin lens at
redshift $z_L$.
\begin{equation}
\begin{aligned}
\frac{ct(\vec\theta)}{(1+z_L)D_{L}}
&= {\textstyle\frac12} |\vec\theta - \vec\beta|^2
   \times \frac{D_{S}}{D_{LS}} \\
&- \frac{4GD_L}{c^2}
   \int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| \, d^2\vec\theta'
\label{full arrival time}
\end{aligned}
\end{equation}
The time here has been scaled to make it dimensionless.  We can make
the density dimensionless as well, by introducing
\begin{equation}
\kappa(\vec\theta) \equiv \frac{\Sigma(\vec\theta)}{\Sigma_L}
\end{equation}
and hence rewrite \eqnref{full arrival time} as
\begin{equation}
\tau(\vec\theta) = {\textstyle\frac12} |\vec\theta - \vec\beta|^2
                   \times \frac{D_{S}}{D_{LS}}
                 - \frac1\pi \int \kappa (\vec\theta')
                   \ln|\vec\theta - \vec\theta'| d^2\vec\theta'
\label{arrival time}
\end{equation}
The scaled arrival time $\tau$ is like a solid angle.  It is of order
the area (in steradians) of the full lensing system --- the expression
$\sim|\vec\theta - \vec\beta|^2$ is of order the image-separation
squared, and the other terms are of similar size.  For this reason, is
convenient to measure $\tau$ in arcsec$^{2}$.

@@

\subsection{Setting up the triaxial N-body mocks}

@@

\begin{table}
\begin{tabular}{llllllll}
Galaxy & $\gamma_\star$ & $M_\star$ & $\gamma_\mathrm{DM}$ & $M_\mathrm{DM}$ & $\Rmap$ \\
\hline
DMdom/DMshallow & 1 & 4 & 0.05 & $11^{2.95}$ & 50 kpc\\ % AA
DMdom/DMcuspy & 1 & 4 & 1 & $11^2$ & 50 kpc \\ % AC
Sdom/DMshallow & 1.5 & $2^{1.5}$ & 0.16 & $11^{2.84}$ & 50 kpc \\ % BB
Sdom/DMcuspy & 1.5 & $2^{1.5}$ & 1 & $11^2$ & 10 kpc %BC
\end{tabular}
\caption{Profile parameters for the four mock galaxies. The name indicates whether 
  the galaxy is centrally dark matter or stellar dominated with a shallow or cuspy 
  dark matter density profile.  Masses are in units of
  $1.8\e{10}\Msun$. The scale lengths for all lenses are
  $(a_\star,a_\mathrm{DM})=(2,20)$\,kpc. $\Rmap$ is the 2D projected radius
used to generate the lens configurations.}
\label{mock galaxy params}
\end{table}

@@

Each of these configurations were modelled with and without time delays and
with and without a central image, for a total of 24 test cases. (The central
image is typically highly demagnified. For galaxy lenses it is very difficult
to find since it lies along the sight line to the bright lensing galaxy; in
clusters, however, such images have been seen (CITES)). We assumed for all our
tests that the lensing mass was radially symmetric (Prior vi):
For our mock data, this is known to be
true; it is most often the case with real galaxies, unless there is an obvious
observed asymmetry. The central pixel was refined into a further nine pixels to
capture any steep rise in the profile.  Two of the four mock galaxies have a
steeply rising inner profile. {\bf JR.  What happens if we don't assume this
symmetry? We ought to add an appendix showing this for one example case.}

@@

The lens equation in its complete form becomes:
%
\begin{eqnarray}
N^2ct(\vec\theta) & = & (1+z_L)\frac{D_{L}D_{S}}{D_{LS}}\frac12 |\vec\theta - \vec\beta|^2 \nonumber \\
& & - (1+z_L)\frac{4GD_{L}^2}{c^2}\int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| d^2\vec\theta'
\label{eqn:full_arrival_time}
\end{eqnarray}
%
where the factor of $D_L^2$ in the second term comes from the fact that $\Sigma$
has units of \Msun/lyr$^2$. We can clean this up by writing down a dimensionless time
delay:
%
\begin{equation}
\tau = \left[ (1+z_L) d_L\right]^{-1}\zeta t
\label{tau}
\end{equation}
%
in terms of our previous definitions. If we
further define a dimensionless density:
%
\begin{equation}
\kappa_\infty = \frac{4\pi G}{c^2}\frac{c}{H_0}d_L\Sigma
              = \frac{d_L}{\zeta}\Sigma
\end{equation}
%
and a lensing potential:
%
\begin{equation}
\psi(\vec\theta) = \frac1\pi \int \kappa_\infty(\vec\theta') \ln|\vec\theta - \vec\theta'| d^2\vec\theta'\
\label{lensing potential}
\end{equation}
%
we can express \eqnref{eqn:full_arrival_time} very compactly as:
%
\begin{equation}
\tau(\vec\theta) = \frac12 \xi |\vec\theta-\vec\beta|^2 - \psi(\vec\theta)
\label{arrival time}
\end{equation}
We explicitly write $\kappa_\infty$ to remind ourselves
that there is no source distance factor involved. This is useful when we
consider multiple sources.

>>>>

In this paper, we use a new program --- \Glass\ --- to determine what
combination of lensing, stellar mass, and stellar kinematic
constraints best constrain the projected mass profile and shape of a
gravitational lens.  The compute-intensive core of \Glass\ is a
parallel implementation of the high-dimensional sampling algorithm
from \cite{2012MNRAS.425.3077L}.  Around that core, \Glass\ has
extensible code (in software terminology, a framework) in Python,
encoding the lens equation, a basis and priors for the mass
distribution, and various post-processing and visualisation modules.  \Glass\
is described in \secref{sec:glass}.  In \secref{sec:theory} before that we
review the elements of lensing theory, stellar population synthesis, and
stellar dynamics we will need.

@@

We follow \cite{1986ApJ...310..568B} with minor differences in notation.

The lens equation
\begin{equation}
\begin{aligned}
\vec\beta &= \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta) \\
\vec\alpha(\vec\theta) &= \frac{4G}{c^2D_L} \int \Sigma(\vec\theta')
                          \frac{(\vec\theta - \vec\theta')}
                          {\ |\vec\theta - \vec\theta'|^2} \, d^2\vec\theta'
\end{aligned}
\label{eqn:lens_equation}
\end{equation}
maps an observed image position $\vec\theta$ to a source position
$\vec\beta$.  The lens can then be thought of as a projected surface
density $\Sigma$ which diverts the path of a photon instantaneously
through the bending angle $\vec\alpha$.  The $D$ factors, as in the previous section, are angular
diameter distances, which depend on the particular choice of cosmological
parameters $\Omega$:
\begin{equation}
D_{LS} = \frac c{H_0} \frac{1+z_S}{1+z_L} \int_{z_L}^{z_S}
                      \frac{dz}{\sqrt{\Omega_m(1+z)^3 + \Omega_\Lambda}}
\end{equation}
One way to understand the lens equation is via Fermat's principle. We
can think of light as travelling only along extremum paths where
lensed images occur. Such paths occur at
the extrema of the photon {\it arrival time\/} $t(\vec\theta)$
that depends on the geometric path the photon takes and the general
relativistic gravitational time dilation due to a thin lens at
redshift $z_L$.
\begin{equation}
\begin{aligned}
\frac{ct(\vec\theta)}{(1+z_L)D_{L}}
&= {\textstyle\frac12} |\vec\theta - \vec\beta|^2
   \times \frac{D_{S}}{D_{LS}} \\
&- \frac{4GD_L}{c^2}
   \int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| \, d^2\vec\theta'
\label{full arrival time}
\end{aligned}
\end{equation}
The time here has been scaled to make it dimensionless.  We can make
the density dimensionless as well, by introducing
\begin{equation}
\kappa(\vec\theta) \equiv \frac{\Sigma(\vec\theta)}{\Sigma_L}
\end{equation}
and hence rewrite \eqnref{full arrival time} as
\begin{equation}
\tau(\vec\theta) = {\textstyle\frac12} |\vec\theta - \vec\beta|^2
                   \times \frac{D_{S}}{D_{LS}}
                 - \frac1\pi \int \kappa (\vec\theta')
                   \ln|\vec\theta - \vec\theta'| d^2\vec\theta'
\label{arrival time}
\end{equation}
The scaled arrival time $\tau$ is like a solid angle.  It is of order
the area (in steradians) of the full lensing system --- the expression
$\sim|\vec\theta - \vec\beta|^2$ is of order the image-separation
squared, and the other terms are of similar size.  For this reason, is
convenient to measure $\tau$ in arcsec$^{2}$.

@@

\subsection{The triaxial N-body mock galaxies}

@@

\begin{table}
\begin{tabular}{llllllll}
Galaxy & $\gamma_\star$ & $M_\star$ & $\gamma_\mathrm{DM}$ & $M_\mathrm{DM}$ & $\Rmap$ \\
\hline
DM-ctr,DM-core & 1 & 4 & 0.05 & $11^{2.95}$ & 50 kpc\\ % AA
DM-ctr,DM-cusp & 1 & 4 & 1 & $11^2$ & 50 kpc \\ % AC
S-ctr,DMcore & 1.5 & $2^{1.5}$ & 0.16 & $11^{2.84}$ & 50 kpc \\ % BB
S-ctr,DMcusp & 1.5 & $2^{1.5}$ & 1 & $11^2$ & 10 kpc %BC
\end{tabular}
\caption{Profile parameters for the four mock galaxies. The name indicates whether 
  the galaxy is centrally dark matter or stellar dominated with a shallow or cuspy 
  dark matter density profile.  Masses are in units of
  $1.8\e{10}\Msun$. The scale lengths for all lenses are
  $(a_\star,a_\mathrm{DM})=(2,20)$\,kpc. $\Rmap$ is the 2D projected radius
used to generate the lens configurations.}
\label{mock galaxy params}
\end{table}

@@

Each of these configurations were modelled with and without time delays and
with and without a central image, for a total of 24 test cases. (The central
image is typically highly demagnified. For galaxy lenses it is very difficult
to find since it lies along the sight line to the bright lensing galaxy; in
clusters, however, such images have been seen (CITES)). We assumed for all our
tests that the lensing mass was radially symmetric (Prior vi):
For our mock data, this is known to be
true; it is most often the case with real galaxies, unless there is an obvious
observed asymmetry. The central pixel was refined into a further 25 pixels to
capture any steep rise in the profile.  Two of the four mock galaxies have a
steeply rising inner profile. {\bf JR.  What happens if we don't assume this
symmetry? We ought to add an appendix showing this for one example case.}

@@

The lens equation in its complete form becomes:
%
\begin{eqnarray}
N^2ct(\vec\theta) & = & (1+z_L)\frac{D_{L}D_{S}}{D_{LS}}\frac12 |\vec\theta - \vec\beta|^2 \nonumber \\
& & - (1+z_L)\frac{4GD_{L}^2}{c^2}\int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| d^2\vec\theta'
\label{eqn:full_arrival_time}
\end{eqnarray}
%
where the factor of $D_L^2$ in the second term comes from the fact that $\Sigma$
has units of \Msun/lyr$^2$. We can clean this up by first writing down a dimensionless time
delay
%
\begin{equation}
\tau = \left[ (1+z_L) d_L\right]^{-1}\zeta t
\label{tau}
\end{equation}
%
in terms of our previous definitions and defining $D_L \equiv (c/H_0)d_L$. We
further define a dimensionless density
%
\begin{equation}
\kappa_\infty = \frac{4\pi G}{c^2}\frac{c}{H_0}d_L\Sigma
              = \frac{d_L}{\zeta}\Sigma
\end{equation}
%
and a lensing potential
%
\begin{equation}
\psi(\vec\theta) = \frac1\pi \int \kappa_\infty(\vec\theta') \ln|\vec\theta - \vec\theta'| d^2\vec\theta'\
\label{lensing potential}
\end{equation}
%
Now we can express \eqnref{eqn:full_arrival_time} very compactly as
%
\begin{equation}
\tau(\vec\theta) = \frac12 \xi |\vec\theta-\vec\beta|^2 - \psi(\vec\theta)
\label{arrival time}
\end{equation}
%
where $\xi=d_S / d_{LS}$.
We explicitly write $\kappa_\infty$ to remind ourselves
that there is no source distance factor involved. This is useful when we
consider multiple sources.
